---
title: "An Overview of the Data Science Workflow using R"
output:
  html_notebook: default
---

# Talk Overview

1. Introduction
2. Overview of RStudio and R notebooks
3. View functionality of Chicago crimes Shiny app
4. Discussion of `tidyverse`, `ggplot2` and other libraries usage 
within source code of app
5. Review how Shiny apps work
6. Questions

# Introduction to R Notebooks

This is an R Notebook. It is an upgrade to R Markdown for the following reasons:

* It is interactive and can be knit to html.
* Single code chunks can be ran one at a time, as well as in line code such as 2 + 2 = `r 2 + 2`. 
* Results are shown without having to re-knit the entire document. 
* Languages other than R can be used.

```{python}
def fib(n):
  a, b = 0, 1
  for _ in xrange(n):
    yield a
    a, b = b, a + b
print list(fib(11))

```

# Workflow Overview

## Load libraries, source functions

Loading the `tidyverse` library loads the following packages:

* ggplot2
* dplyr
* tidyr
* readr
* purrr
* tibble

```{r}
library(tidyverse)
library(lubridate)
library(stringr)
source("chicago_crime/source_functions.R")
```

## Read and prepare data

The `readr` library can be useful for reading a variety of file formats and can be much faster than base R data reading functions. 

```{r, eval=FALSE}
crime_dat_raw <- readr::read_csv("chicago_crime.csv")
crime_dat <- edit_raw_data(raw_data = crime_dat_raw)

saveRDS(crime_dat, "chicago_crime/crime_dat_2016.RDS")
```

```{r}
crime_dat <- readRDS("chicago_crime/crime_dat_2016.RDS")
```


## Manipulating data

We can access a function stored in `source_functions` to cleanly filter the data based on user inputs. 

```{r}
crime_filter <- filter_crime_data(crime_dat = crime_dat, 
                                  crime = "Burglary", 
                                  start = "2016-01-01", 
                                  end = "2016-02-01", 
                                  group = "Description")
```

### dplyr

#### Pipe operator: `%>%`

The pipe operator (originally part of the `magrittr` library)  makes for very readable, smart code. 

The pipe operator works by taking an object and piping it to the first argument of the next function. 

##### Poem Example

Below is an example from Hadley Wickham's example of the enhanced readability of code when using the pipe operator over standard base R functionality.

> Little bunny Foo Foo
Hopping through the forest
Scooping up the field mice
And bopping them on the head

Using base R functionality:
```{r, eval = FALSE}
foo_foo <- little_bunny()
bop_on(scoop_up(hop_through(foo_foo, forest), field_mouse), head)
```

Using tidyverse functionality:
```{r, eval = FALSE}
foo_foo %>%
  hop_through(forest) %>%
  scoop_up(field_mouse) %>%
  bop_on(head)
```

#### Non-standard evaluation

Non-standard evaluation is a unique syntactical feature within R that allows the program to use the name of an object in addition to its value. 

Non-standard evaluation is used for three common purposes:

1. Labelling -- uses the expressions rather than the values as labels
2. Non-standard scoping -- looks for objects in a data frame or list before looking in the global environment
3. Metaprogramming -- computing on unevaluated code


```{r}
num_seq <- 1:10
num_seq_squared <- num_seq^2

plot(num_seq, num_seq_squared)
```

## Visualizing data

### ggplot2

The `ggplot2` library is based on the "grammar of graphics", a system of generating graphs based on a few simple components:

* a single or multiple data sets
* geometries (geom) which are visual representations of data points with corresponding aesthetics (aes) such as x, y, size, color
* a coordinate system

Aesthetics are mapped from the data frame to the geom layer to create a simple plot that can be customized (scale, colors, window, theme, etc.) with additional elements. 

#### Plot data preparation 

```{r}
crime_burg <- filter_crime_data(crime_dat = crime_dat, 
                                     start = "2016-01-01", 
                                     end = "2016-12-31", 
                                     crime = "Burglary", 
                                     group = "Description")

barplot_data_list <- create_barplot_data(crime_filter = crime_burg, 
                                    time_measure = "hour", 
                                    north = 41.96766,
                                    east = -87.37907,
                                    south = 41.78872,
                                    west = -87.88033)

```

Barplot using `ggplot2`:

```{r}
head(long_data)
```


```{r}
plot_barplot_by_time_loc(barplot_data_list)
```

Using base R plotting functionality:

```{r}
head(barplot_data_list$table_data)
```

```{r}
barplot(height = t(barplot_data_list$table_data), 
        col = barplot_data_list$colors,
        xlab = "Hour", ylab = "Total", 
        legend = colnames(barplot_data_list$table_data))
```

## Interactive Shiny app

Shiny is a library developed by RStudio to create easy, interactive web applications without the use of JavaScript. 

A basic shiny app consists of:

* A UI object
* A server function

```{r}
options(shiny.reactlog = TRUE)
shiny::showReactLog()
```

# References

https://catalog.data.gov/dataset/crimes-2001-to-present-398a4

http://adv-r.had.co.nz/Computing-on-the-language.html

https://shiny.rstudio.com/articles/reactivity-overview.html 

https://shiny.rstudio.com/reference/shiny/latest/showReactLog.html

